Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2022-12-28 paper-1/manuscript.tex
Rev sec Results:F2

% Our first aim was to replicate deviance detection properties reported in dissociated neuronal cell culture by \cite{Kubota2021-dx}, reproduced in Figure \ref{fig2}a. In their case, cell cultures -- like our model, randomly connected networks with two-dimensional extent -- were stimulated at 10 different sites in msc sequences, and at two sites with 10\%/90\% probability in dev and std sequences. Deviance detection was assessed analogously to our definition of DDI and reached an average level of around 0.3 across two cultures.

% In our model, we could readily identify networks and stimuli whose responses developed in a qualitatively similar fashion, with high activity in response to target in the dev sequence, low activity in the std sequence, and intermediate activity in the msc sequence, as illustrated with an example in Figure \ref{fig2}b. The model responses developed on a compressed time scale compared to neuronal culture, which we tentatively attribute to unfaithful aspects of the model, including a lack of conduction delay or spontaneous activity, a much smaller number of participating neurons, and structural differences.

% Across networks and stimuli (Figure \ref{fig2}c), we found that our model reliably produced positive mismatch responses (T = 6.67e+03, p = 9.26e-16; one-sided Wilcoxon signed-rank test, n = 120) while both SSA index (T = 6.19e+03, p = 9.76e-12) and DDI (T = 5.97e+03, p = 4.75e-10) were also positive. The average DDI value reached around 0.1, which was less than the values reported by \cite{Kubota2021-dx}, but note that our target stimuli were presented more frequently (20\%) than theirs (10\%) in the dev and msc sequences.

>>>>

To show that our model is capable of deviance detection, we first show the responses of the sample network in all conditions (dev, msc, and std sequences) in Figure \ref{fig2}a. Here, each row corresponds to one trial, aligned to stimulus input at t = 0, with the number of spikes in each millisecond bin indicated by color. All responses share an early activity pattern, which derives from the stimulated neurons and their immediate postsynaptic targets, but differences soon start to emerge. The dev sequence responses, shown at the top of the plot, exhibit a clear peak in activity around 15 ms after stimulation. The std sequence responses show a later, more diffuse activity peak around 20 ms, and the msc sequence responses lack any obvious activity cluster. We sum these patterns up in Figure \ref{fig2}b, where we show the mean network activity, in terms of spikes per millisecond bin, across trials of each type. Clearly, the target response in the dev sequence is greater than either the msc or the std response.

We could readily identify similar patterns emerging across most networks and stimulus assignments. To show that deviance detection is robust to network and stimulus identity, we calculated the DDI and related indices mentioned above for each network, and show them in a summary plot in Figure \ref{fig2}c. We found that our model reliably produced positive mismatch indices (T = 6.67e+03, p = 9.26e-16, median 0.253; one-sided Wilcoxon signed-rank test, n = 120) while both SSA index (T = 6.19e+03, p = 9.76e-12, median 0.135) and DDI (T = 5.97e+03, p = 4.75e-10, median 0.114) were also positive, despite some outliers. This shows that our model was well behaved and suitable for an investigation of deviance detection.
